\begin{abstract}

  Computational semigroup theory is an area of research that is subject to
  growing interest.  The development of semigroup algorithms allows for new
  theoretical results to be discovered, which in turn informs the creation of
  yet more algorithms.  Groups have benefitted from this cycle since before the
  invention of electronic computers, and the popularity of computational group
  theory has resulted in a rich and detailed literature.  Computational
  semigroup theory is a less developed field, but recent work has resulted in a
  variety of algorithms, and some important pieces of software such as the
  \Semigroups{} package for \GAP{}.

  Congruences are an important part of semigroup theory.  A semigroup's
  congruences determine its homomorphic images in a manner analogous to a
  group's normal subgroups.  Prior to the work described here, there existed few
  practical algorithms for computing with semigroup congruences.  However, a
  number of results about alternative representations for congruences, as well
  as existing algorithms that can be borrowed from group theory, make
  congruences a fertile area for improvement.  In this thesis, we first consider
  computational techniques that can be applied to the study of congruences, and
  then present some results that have been produced or precipitated by applying
  these techniques to interesting examples.

  After some preliminary theory, we present a new parallel approach to computing
  with congruences specified by generating pairs.  We then consider alternative
  ways of representing a congruence, using intermediate objects such as linked
  triples.  We also present an algorithm for computing the entire congruence
  lattice of a finite semigroup.  In the second part of the thesis, we classify
  the congruences of several monoids of bipartitions, as well as the principal
  factors of several monoids of partial transformations.  Finally, we consider
  how many congruences a finite semigroup can have, and examine those on
  semigroups with up to seven elements.
  
  \thispagestyle{plain}
\end{abstract}
