\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

Attempting to complete a PhD has been a great undertaking, and in completing
this thesis I am nearing the end of an important chapter in my life.  The years
I have spent as a postgraduate researcher have probably been the happiest of my
life, but at times the work involved has been tough, and without the support of
people around me I certainly couldn't have made it this far.  Almost everyone I
have met and got to know during this period has touched my life in a positive
way, but there are a few people in particular that I wish to thank.

Firstly, I would like to thank my supervisor James D.~Mitchell, for his honesty
and friendliness, and for the many hours he has spent correcting my work and
making me a better mathematician.  Secondly, I would like to thank my friend and
office-mate Wilf Wilson, whose wonderful company has kept me from falling asleep
through many weary afternoons of writing and coding.  I am also indebted to the
Engineering and Physical Sciences Research Council (EPSRC), whose generous grant
has allowed me to pursue computational semigroup theory freely for the last four
years.

Finally, I wish to thank Claire Young.  She has been the most important part of
my life throughout my postgraduate career, and her love and support during the
tougher moments of this PhD have given me the motivation to overcome what
sometimes felt like insurmountable obstacles.

\begin{flushright}
  Michael Torpey

  \singlespacing
  \textit{St Andrews \\July 2018}
\end{flushright}

% TODO: Further acknowledgements
