\newgeometry{tmargin=2.25cm, lmargin=2.25cm, rmargin=2.25cm, bmargin=3.25cm}

\noindent{\Huge \textbf{Declarations}}

\vspace{-1.3em}
\section*{Candidate's declaration}

\noindent
I, Michael Colin Torpey, do hereby certify that this thesis, submitted for the
degree of PhD, which is approximately
% texcount -1 -inc declarations.tex abstract.tex && texcount -1 -inc thesis.tex
59,000
words in length, has been written by me, and that it is the record of work
carried out by me, or principally by myself in collaboration with others as
acknowledged, and that it has not been submitted in any previous application for
any degree.
\\

\noindent
I was admitted as a research student at the University of St Andrews in
September 2014.
\\

\noindent
I received funding from an organisation or institution and have acknowledged the
funder(s) in the full text of my thesis.
\\

\vspace{1.0em}
\noindent
Date:\makebox[7em]{\dotfill}
Signature of candidate:\dotfill
\\

\vspace{-1.3em}
\section*{Supervisor's declaration}

\noindent
I hereby certify that the candidate has fulfilled the conditions of the
Resolution and Regulations appropriate for the degree of PhD in the University
of St Andrews and that the candidate is qualified to submit this thesis in
application for that degree.
\\

\vspace{1.0em}
\noindent
Date:\makebox[7em]{\dotfill}
Signature of supervisor:\dotfill
\\

\vspace{-1.3em}
\section*{Permission for publication}

\noindent
In submitting this thesis to the University of St Andrews we understand that we
are giving permission for it to be made available for use in accordance with the
regulations of the University Library for the time being in force, subject to
any copyright vested in the work not being affected thereby. We also understand,
unless exempt by an award of an embargo as requested below, that the title and
the abstract will be published, and that a copy of the work may be made and
supplied to any bona fide library or research worker, that this thesis will be
electronically accessible for personal or research use and that the library has
the right to migrate this thesis into new electronic forms as required to ensure
continued access to the thesis.
\\

\noindent
I, Michael Colin Torpey, confirm that my thesis does not contain any third-party
material that requires copyright clearance.
\\

\noindent
The following is an agreed request by candidate and supervisor regarding the
publication of this thesis: no embargo on print copy; no embargo on electronic
copy.
\\

\vspace{1.0em}
\noindent
Date:\makebox[7em]{\dotfill}
Signature of candidate:\dotfill
\\

\vspace{1.0em}
\noindent
Date:\makebox[7em]{\dotfill}
Signature of supervisor:\dotfill
\\

\vspace{-1.3em}
\section*{Underpinning Research Data or Digital Outputs}

%\subsection*{Candidate's declaration}

I, Michael Colin Torpey, hereby certify that no requirements to deposit original
research data or digital outputs apply to this thesis and that, where
appropriate, secondary data used have been referenced in the full text of my
thesis.
\\

\vspace{1.0em}
\noindent
Date:\makebox[7em]{\dotfill}
Signature of candidate:\dotfill
\\

\restoregeometry
